---
title: "New Manager Bounce"
subtitle: "Data Preparation"
author: "Josh Muncke"
date: "20/03/2019"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation

Step 2 is to cleanse, prepare and join the manager and game data.

## Setup 

Load the required R packages.

```{r load our packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(feather)
library(lubridate)
library(testthat)
library(redbull)
```

Load the raw extracted data into memory.

```{r load raw data}
games_raw <- read_feather("../data/raw/game_data_raw.feather")
managers_raw <- read_feather("../data/raw/manager_data_raw.feather")
club_names <- read_feather("../data/raw/mapping_data_raw.feather")
```

## Game data

### Replace club names

The names of teams between the manager and game data is slightly different for some clubs (e.g. "Man Utd" vs "Manchester United"). In order to remedy this we will use the mapping table we created in the first step to replace the team names for both the home and away team columns of each game.

```{r map club names}
games_prepared <- games_raw

games_prepared %<>% 
  left_join(club_names, by = c("HomeTeam" = "short_name")) %>%
  mutate(HomeTeam = ifelse(is.na(long_name), HomeTeam, long_name)) %>%
  select(-long_name) %>% 
  left_join(club_names, by = c("AwayTeam" = "short_name")) %>%
  mutate(AwayTeam = ifelse(is.na(long_name), AwayTeam, long_name)) %>%
  select(-long_name)
```

### Expand game data

The raw game data has one row per match with the results for both the home and away teams. Since we want to do our analysis at a club-by-club level we effectively need to double this data so that we have a single row per team per game indicating who the opponent was, the scores for each and whether that team was playing at home or away.

We first add an identifier column (just the row number) for each game so we can keep track of missing/duplicates later on.

```{r add game id}
games_prepared %<>% 
  select(game_date = Date, everything()) %>% 
  mutate(game_id = row_number())
```

We now update the game date column to format it to an actual date (there's a couple of different date formats to work with).

```{r update game date}
games_prepared %<>% 
  mutate(game_date = case_when(
  str_length(game_date) == 10 ~ as_date(game_date, format = "%d/%m/%Y", tz = "UTC"),
  str_length(game_date) == 8 ~ as_date(game_date, format = "%d/%m/%y", tz = "UTC"),
  TRUE ~ as_date('1900-01-01', tz = "UTC")))
```

And finally we create two temporary versions of this data - one for the home result and one for the away result - before binding them together.

We also make sure to create a "points" column indicating the number of points the team received for the match (0 for a loss, 1 for a draw, 3 for a win).

```{r create results data}
home_games <- games_prepared %>% 
  select(game_id, game_date, game_team = HomeTeam, goals = FTHG, opponent = AwayTeam, opponent_goals = FTAG) %>%
  mutate(game_location = "Home", 
         points = case_when(goals > opponent_goals ~ 3, goals < opponent_goals ~ 0, TRUE ~ 1))

away_games <- games_prepared %>% 
  select(game_id, game_date, game_team = AwayTeam, goals = FTAG, opponent = HomeTeam, opponent_goals = FTHG) %>%
  mutate(game_location = "Away", 
         points = case_when(goals > opponent_goals ~ 3, goals < opponent_goals ~ 0, TRUE ~ 1))

game_results_full <- bind_rows(home_games, away_games)

rm(list = c('home_games','away_games'))
```

### Test and visualize

A good test would be to make sure we have the right number of expected games between the `full` and `prepared` data sets and also check we don't have any games that have greater or less than two records.

We're going to use the `testthat` package to check this.

```{r compare game ids}
test_that('Same number of distinct game ids',
          expect_equal(
            games_prepared %>% distinct(game_id) %>% tally(),
            game_results_full %>% distinct(game_id) %>% tally()
          ))
```

```{r check only two records per game}
test_that('Same number of distinct game ids',
          expect_equal(
            game_results_full %>% add_count(game_id) %>% filter(n != 2) %>% tally() %>% pull(nn),
            0
          ))
```

Finally let's do a quick analysis to look at the average number of games per month over time - just a simple visual gut check. I'm going to use `ggplot2` and shamelessly self-promote the `[redbull](http://github.com/deathbydata/redbull)` package I developed for themeing.

```{r visualize games over time}
theme_set(theme_rb())

game_results_full %>%
  mutate(game_month = round_date(game_date, "month")) %>%
  group_by(game_month) %>%
  summarize(n_games = n_distinct(game_id)) %>%
  ggplot(aes(x = game_month, y = n_games)) +
  geom_line() +
  labs(title = "Number of EPL games per month since 1995",
       caption = "Source: football-data.co.uk",
       x = "Month",
       y = "Number of games")
```

All looks good so we write this prepared game data back locally.

```{r save game data}
game_results_full %>%
  write_feather("../data/prep/game_data_prep.feather")
```

```{r remove un-needed dfs}
rm(list = c('club_names', 'games_prepared', 'games_raw'))
```

## Manager data

Like most data that is scraped from a web-page the manager data is a little ugly and requires some love.

For our purposes we only require the name of the manager, the club they managed and the time window so we `select` these columns and drop the rest.

```{r prepare manager data}
managers_prepared <- managers_raw

managers_prepared %<>% select(manager_name = Name, 
                          managed_club = Club, 
                          managed_from = From, 
                          managed_until = Until)
```

Some of the date data needs to be fixed. In particular for incumbent managers the `managed_until` column has the value "Present*" which we will replace with today's date.

We also coerce the string dates into a formatted date field.

```{r fix manager dates}
managers_prepared %<>% 
  mutate(managed_until = if_else(managed_until == 'Present*', format.Date(Sys.Date() + 1, format = "%e %B %Y"), managed_until))

managers_prepared %<>% mutate(managed_from = lubridate::as_date(managed_from, format = "%e %B %Y", tz = "UTC"), 
                          managed_until = lubridate::as_date(managed_until, format = "%e %B %Y", tz = "UTC"))
```

Need to fix a few little errors in this data.

For example there are some clubs which have two managers simultaneously (e.g. Ray Lewington and Ron Noades at Crystal Palace in 1998). Whilst that's logically okay it doesn't really work for this analysis since it will causes duplicate results.

```{r manager cleanup}
managers_prep %<>%
  filter(! (manager_name == "Tomas Brolin" & managed_club == "Crystal Palace")) %>%
  filter(! (manager_name == "Ray Lewington" & managed_club == "Crystal Palace")) %>%
  filter(! (manager_name == "Mike Stowell" & managed_club == "Leicester City")) %>%
  filter(! (manager_name == "Kevin Bond" & managed_club == "Queens Park Rangers")) %>%
  mutate(managed_until = if_else((manager_name == "Roy Evans" & managed_club == "Liverpool" & managed_until == as.Date("1998-11-12")), as.Date("1998-06-30"), managed_until))
```

Also - some of the management date ranges are a little tricky. In some cases the new manager takes over the day *after* the old manager (this is fine). In other cases it's the *same* day (this causes problems). In yet other cases it is many days later (could also causes problems).

This all requires manual investigation and fixing.

```{r fix manager date ranges}
managers_prep %<>%
  arrange(managed_club, managed_from) %>%
  group_by(managed_club) %>%
  mutate(next_mgr_started = lead(managed_from)) %>% 
  mutate(overlap = as.numeric(next_mgr_started - managed_until)) %>%
  mutate(overlap = if_else(is.na(overlap), 1, overlap)) %>%
  mutate(managed_until = if_else(overlap == 0, managed_until - 1, managed_until)) %>%
  dplyr::select(-next_mgr_started, -overlap) %>%
  ungroup()
```


### Fixing game data



Let's also dump this to a feather.

```{r write game results full}
game_results_full %>% write_feather("../data/game_results_full.feather")
```

## Fix results data

Load back in our feather file.

```{r load results data}
game_results_full_raw <- read_feather("../data/game_results_full.feather")
managers <- read_feather("../data/managers.feather")
```

Check that we have no missing teams in our game data (i.e. every row has a manager)

```{r check no missing teams}
library(testthat)

test_that('All games have a manager', expect_equal(game_results_full_raw %>% 
  anti_join(managers,  by = c("game_team" = "managed_club")) %>%
  select(game_team) %>%
  distinct() %>%
  nrow(), 0))
```

## Join results to managers by club name

Use the club name to join results to managers. We also create a "valid_date" column which tracks whether the game date is between the management dates for the manager.

```{r join results}
full_result_data_raw <- game_results_full_raw %>%
  left_join(managers, by = c("game_team" = "managed_club")) %>%
  mutate(valid_date = (game_date >= managed_from & game_date <= managed_until))
```

Every game id should appear twice, once for the home and once for the away side. How many games appear less than or more than twice?

```{r examine invalid games}
full_result_data_raw %>% 
  filter(valid_date) %>%
  add_count(game_id) %>% 
  filter(n != 2) %>%
  distinct(game_id) %>%
  count()

full_result_data_raw %>% 
  filter(valid_date) %>%
  add_count(game_id) %>% 
  filter(n != 2) %>%
  group_by(game_id, game_team) %>%
  mutate(team_n = n()) %>%
  ungroup() %>%
  arrange(game_id) %>%
  select(game_id:opponent_goals, manager_name:managed_until, n:team_n) %>%
  filter(n == 1 | team_n > 1) %>%
  View()
```

Let's have a deeper look at these 24 games...
```{r}
full_result_data_raw %>% 
  filter(valid_date) %>%
  add_count(game_id) %>% 
  filter(n != 2) %>%
  arrange(game_id) %>%
  knitr::kable()
```

We see a lot of them are due to overlapping management periods. 

We will manually go back and fix these (one day). For now we're going to just remove these 24 games from the dataset.

```{r filter valid games}
full_result_data_valid <- full_result_data_raw %>%
  filter(valid_date) %>%
  add_count(game_id) %>% 
  filter(n == 2)
```

Write to feather!

```{r}
full_result_data_valid %>% write_feather("../data/game_results_valid_final.feather")
```